---
title: "TreeSE_Coxmodel.qmd"
format: html
editor: visual
---

# I-Fitting a Cox proportional hazard model into all- pairwise log-ratio data format

Load libraries

```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(mia)
library(ggplot2)
library(lattice)
library(TreeSummarizedExperiment)
library(coda4microbiome)
library(SummarizedExperiment)
library(tibble)
library(phyloseq)
```

read the data set

```{r}
load("~/work/biom.Rdata")
```

convert phyloseq object to TreeSE

```{r}
biom_TSE <- convertFromPhyloseq(biom)
```

split the data into training and testing subsets

```{r}
# Set seed for reproducibility
set.seed(123)
# Split TreeSE into training and test sets by samples (columns)
num_samples <- ncol(biom_TSE)
train_indices <- sample(seq_len(num_samples), size = floor(0.7 * num_samples))
test_indices <- setdiff(seq_len(num_samples), train_indices)

# Subset TreeSE
biom_TSE_train <- biom_TSE[, train_indices]
biom_TSE_test <- biom_TSE[, test_indices]
```

Prepare data for Cox model training

```{r}
# Extract colData
col_data_train <- as.data.frame(colData(biom_TSE_train))
# Extract time-to-event and status from colData
time <- col_data_train$T1Dweek   # numeric vector
status <- col_data_train$T1D  # binary vector (0 = censored, 1 = event)
covars <- col_data_train[, c("Antibiotics", "Sex")]  # update with correct column names
# Extract abundance matrix (assumes 1st assay)
x <- t(assay(biom_TSE_train))  # transpose: rows = samples, columns = features
# Replace row names with only digits extracted from original row names
rownames(x) <- gsub("\\D", "", rownames(x))
colnames(x) <- gsub("\\D", "", colnames(x))
```

Train the Model

```{r}
coxmodel_logratio <- coda_coxnet(
  as.matrix(x),
  time = time,
  status = status,
  covar = covars,
  lambda = "lambda.1se",
  alpha = 0.9,
  nfolds = 5,
  showPlots = FALSE,
  coef_threshold = 0.01
)
```

output trained Cox model

```{r, message=FALSE, warning=FALSE, echo=TRUE}
print(coxmodel_logratio)
```

plot survival curve

```{r}
survcurve <- plot_survcurves(risk.score = coxmodel_logratio$risk.score,
                             time,
                             status,
                             strata.quantile = 0.5)
survcurve
```

Prepare the testing data

```{r}
#prepare the testing subset
col_data_test <- as.data.frame(colData(biom_TSE_test))
# Extract time-to-event and status from colData
time_test <- col_data_test$T1Dweek   # numeric vector
status_test <- col_data_test$T1D  # binary vector (0 = censored, 1 = event)
covars_test <- col_data_test[, c("Antibiotics", "Sex")]  # update with correct column names
# Extract abundance matrix (assumes 1st assay)
x_test <- t(assay(biom_TSE_test))  # transpose: rows = samples, columns = features
# Replace row names with only digits extracted from original row names
rownames(x_test) <- gsub("\\D", "", rownames(x_test))
colnames(x_test) <- gsub("\\D", "", colnames(x_test))
```

calculate the Risk Score

```{r}
# Ensure coef_vec is numeric and named
coef_vec <- as.numeric(coef(coxmodel_logratio))
names(coef_vec) <- names(coef(coxmodel_logratio))

# Make sure x_test is a numeric matrix
xt_tes <- as.matrix(x_test)

# Align x_test columns to match the coefficient names
x_test <- x_test[, names(coef_vec), drop = FALSE]

# Compute the risk scores
risk_score_test <- as.vector(x_test %*% coef_vec)
```

Calculates Harrell's C-index:

```{r}
library(survival)
concordance <- survConcordance(Surv(time_test, status_test) ~ risk_score_test)
print(concordance$concordance)
```
